Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2022-01-10 analysis/paper/paper.Rmd
Revisions and replies to reviewers

However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [@barton2011; @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1977; @binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvination, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.

@@

As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns [e.g. @Bicho2020; @kitchel2021] and are of central importance in Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000]. The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots.

>>>>

However, for the most part we lack even a most basic understanding of what any individual lithic object in an assemblage has been used for [@dibble2017]. For example, a vast amount of artefacts defined as debitage are likely to have fulfilled the function of tools, and both debitage and formal tool types could have had various different purposes and had a multitude of shapes throughout their use-life. This has major implications that the above-outlined analysis does
not take properly into account, rendering it difficult to align any identified pattern with specific behavioural dimensions. As a consequence, the second part of the analysis employs a suite of measures developed for the classification of lithic assemblages with these inferential limitations in mind [@barton2011; @clark2017, and below]. The logic behind these measures are founded on an understanding of technology as being organised along a continuum ranging between curated and expedient [@binford1973; @binford1977; @binford1979]. An expedient technological organisation pertains to the situational production of tools to meet immediate needs, with little investment of time and resources in modification and rejuvenation, resulting in high rates of tool replacement. Curated technological organisation, on the other hand, has been related to manufacture and maintenance of tools in anticipation of future use, the transport of these artefacts between places of use, and the modification and rejuvenation of artefacts for different and changing situations.

@@

As these measures are argued to predominantly be determined by land-use and mobility patterns, relative frequency of chips and relative frequency of non-flint material are also included in the analysis, as these measures have also been linked to mobility patterns [e.g. @Bicho2020; @kitchel2021] and are of central importance to Norwegian Stone Age archaeology [e.g. @breivik2016a; @reitan2016]---the use of local non-flint material has been taken to indicate reduced mobility and increased familiarity with local surroundings [@jaksland2001, 112; @glørstad2010, 181].

@@

The general impression from the CA is that a chronological dimension accounts for a substantial amount of patterning in the data (Figure \@ref(fig:cor)). This is indicated by the general transition across the colour scale in the row plot (Figure \@ref(fig:cor)A), as well as the horseshoe curve or Guttman effect evident in the column plot [Figure \@ref(fig:cor)B, @baxter1994, 119--120; @lockyear2000]. The fact that the two first dimensions of the CA accounts for as much as `r round(explained_cumul_dim2, 2)`% of the inertia or variance also means that the structure of the data is well-represented in the plots and that these therefore are likely reflect true patterning in the data.
